Rotation effects on the Lyman-$\alpha$ line morphology in distant galaxies

### 2013-06-05 paper/Analisis.tex
some new paragraphs

\title[Rotation in the Lyman-$\alpha$ line]{Rotation effects on the
  Lyman-$\alpha$ line morphology in distant galaxies}
\author[N. Garavito and J.E. Forero-Romero]{
\parbox[t]{\textwidth}{\raggedright 
  Nicolas Garavito-Camargo$^{1}$,
  Jaime E. Forero-Romero$^{1}$ 
}
\vspace*{6pt}\\
$^{1}$Departamento de F\'{i}sica, Universidad de los Andes, Cra. 1
No. 18A-10, Edificio Ip, Bogot\'a, Colombia 
}
\maketitle

\begin{abstract}
Rotation is present in the gas kinematics of galaxies up to the
highest redshifts. In this paper we present for the first time
radiative transfer calculations that show the impact of rotation on
the morphology of the Lyman $\alpha$ line. To this end we construct
simplified models where a galaxy is modeled as an homogeneous sphere
composed as an homogeneous mixture of dust and hydrogen at a constant
temperature. These spheres have a solid-body rotation with linear
velocities at the surface in the range $0-300$ \kms. We consider
radiation sources both in the center of the rotating cloud and also
homogeneously distributed around the sphere. We find that higher
rotational velocities increase the width of each peak in the outgoing
line profile while it also increases the amount of Lyman alpha photons
escaping in the line center. This trends makes that for high
rotational velocities and large Hydrogen optical depths the double
peak of the line tends to be erased an be replaced by a single peak the
lines center. This is more pronounced for radiation sources
homogeneously distributed. Concerning the escape fraction we find that
rotation does not have any effect, provided that all the sources are
centrally emitted. However in the case of homogeneously emitted
sources we measure an increase of about a factor of $2$ in the escape
fraction for higher rotational velocity values. 
Our work shows clearly that gas rotation has a non
negligible impact on the shape of the Lyman $\alpha$ line.

\end{abstract}
\begin{keywords}
galaxies: high-redshift - galaxies: star formation - line: formation
\end{keywords}

@@

Due to the resonant nature of the Lyman alpha line, gas kinematics
play an important role shaping its morphology. In the literature there
has been extensive studies of outflow/inflow configurations.

In this paper we study for the first time the impact of rotation on
the morphology of the Lyman $\alpha$ line. To isolate the effects of
rotation we focus on a simple system: the gas distribution is
spherical, with homogeneous density and the gas rotates as a solid
body. We base our work on the Monte Carlo base radiative transfer code
CLARA \citep{CLARA}.

@@

\section{Implementation of Bulk Gas Rotation}
\label{sec:implementation}

We implement into CLARA the simplest model whereby a sphere rotates
with homogeneous angular velocity. We define a cartesian coordinate
system with its origin at the center of the sphere and the rotation
axis to be the $z$-axis, the components in the bulk velocity field, $\vec{v}
= v_{x}\hat{i} + v_{y}\hat{j} + v_{z}\hat{k}$. in the gas can be written as

@@

\section{Grid of Simulated Models}
\label{sec:models}

@@

%poner las graficas para un modelo dado poner el efecto de
%la rotación (poner %uno con velocidad cero) y del observador.

@@

%\begin{figure*}%QUITARLA Y COMENTARLA
  %  \includegraphics[width=0.45\textwidth]{WidthvsThetaCentral.png}
   % \includegraphics[width=0.45\textwidth]{WidthvsTheta.png}
    %\caption{Width of the lyman-alpha line for all the models. 
 % \label{figure:widthvstheta}
%}
%\end{figure*}

@@

%\begin{figure*}
%    \includegraphics[width=0.45\textwidth]{maximumvscosthetaHOMSameSpeed.png}
%    \includegraphics[width=0.45\textwidth]{maximumvscosthetaDifSpeedCentral.png}
%\caption{Position of the maxima
%    in the outgoing spectra for differents Rotational velocities,
%    (up)Central Distriution, (Down) Homogeneous Distribution.%\label{fig:maximumsvsvelocity}} 
%\end{figure*}

%\begin{figure}
%    \includegraphics[width=0.45\textwidth]{maximumvsODDifSpeedsCentral.png}
%    \includegraphics[width=0.45\textwidth]{maximumvsODDifSpeedsHOM.png}
%  \label{fig:maximavsOD}\caption{Position of the maxima in the
%    outgoing spectra for differents Optical Depths, (up)Central
%    Distriution, (Down) Homogeneous Distribution.} 
%\end{figure}

@@

%In Figure ~\ref{fig:maximavsangle} we present the position of the
%maxima as a function of the viewing angle $\mu$ for all the optical
%depths and rotational velocities and the two different photon
%distributions [FALTA LA GRAFICA].

%Finally when we take into account the viewing angle Fig
%~\ref{fig:DifferentObservers} we found that $X_{m}$ does not depend on
%it, but again there is a threshold for the homogeneous distribution in
%which  the double peaked merged into a single peak but this time it is
%just the angle of viewing effect.

%rotational velocities photons escape with
%less scatterings (fig:ref), for $v=100$\kms the majority of the photons
%still escape with scatterings (double-peaked line) but for $v=200$\kms
%the number of photons that escape with a few or any scatterings is higher
%(single-peaked line), this is one of the main effects of rotation. \\

% try a Nscatt plot as a function of theta
%explain the method

% explicar como se mide el ancho para algunos casos
% I dont understand this at all, hope Nscatts plot helps 
 %make comparison with Verhamme et al

@@

%\begin{figure}
 %   \includegraphics[width=0.54\textwidth]{test4.png}
  %  \includegraphics[width=0.54\textwidth]{test4.png}%this is the central}

%\caption{Histogram of $N_{scatt}$ vs $X_{m}$ for the central distribution, The upper plot is for $V_{max}=50km/s $, the bottom plot is for $V_{max}=300km/s$ \label{fig:NscattHisto}} 
%\end{figure}

@@

%\begin{figure}%QUITARLA Y COMENTARLA
 % \includegraphics[width=0.45\textwidth]{FEHomogeneousvsThetaL.png}
 %\label{figure:efvstheta}\caption{Escape fraction dependency with the viewing angle in the %homogeneous configuration.} 
%\end{figure}

@@

\subsection{Inhomogeneous model}

@@

\begin{table*}
\begin{center}
\begin{tabular}{cccc}\hline
Model & SSA22a 4980/80   & HS1549 4667/80 & HS1700 4018/90\\
\hline
V0HOM & 0.99 / 0.90 / 0.40 & 1.0 / 0.96 / 0.52 & 1.0 / 0.99 / 0.66 \\
V100HOM & 0.94 / 0.81 / 0.40 & 0.98 / 0.90 / 0.51 & 0.99 / 0.96 / 0.64 \\
V200HOM & 0.74 / 0.74 / 0.40 & 0.81 / 0.81 / 0.48 & 0.87 / 0.87 / 0.59 \\
V300HOM & 0.59 / 0.59 / 0.39 & 0.66 / 0.66 / 0.45 & 0.73 / 0.73 / 0.54 \\
V0HOMDust & 0.99 / 0.99 / 0.93 & 0.99 / 0.99 / 0.86 & 0.99 / 0.99 / 0.92\\
V100HOMDust & 0.99 / 0.98 / 0.91 & 0.98 / 0.96 / 0.86 & 0.99 / 0.98 / 0.9\\
V200HOMDust & 0.91 / 0.90 / 0.86 & 0.83 / 0.81 / 0.72 & 0.88 / 0.87 / 0.81\\
V300HOMDust & 0.83 / 0.82 / 0.79 & 0.67 / 0.66 / 0.61 & 0.74 / 0.74 / 0.70\\
\hline
\end{tabular}
\caption{
Values of fluxes for tree different NB and the our models simulated with CLARA
} 
\label{table:NBfilters}
\end{center}
\end{table*}

@@

Concerning the escape fraction under rotational effects on the Ly$\alpha$ 
emission line, we found that the escape fraction increase in about $20\%-30\%$
for the homogeneous sphere model. While rotational effects are negligible for
the central models and the escape fraction remains constant. Also the observer
viewing angle have no effect in the escape fraction neither for the homogeneous
and central models. Complementing this analysis we study the average number of 
scatterings $<N_{scatt}>$ that photons perform before escaping of the cloud
taking into account rotational effects. The main result here is for the homogeneous
models for which as velocity increase photons escape with about $\sim 39\%$ 
less scatterings than in the static case.

@@

% fix this table is too big
%\begin{table*}
%\begin{center}
%\begin{tabular}{cc}\hline
%Model & Fit\\
%Central, $\tau=10^5, V=200$\kms &$FWHM(\mu)=-149877+84509\mu-17864\mu^{2}-1677\mu^{3}-59\mu^{4}$\\
%Central, $\tau=10^6, V=200$\kms &$FWHM(\mu)=-32264+9644\mu-1080\mu^{2}+53.8\mu^{3}-\mu^{4}$\\
%Central, $\tau=10^7, V=200$\kms & $FWHM(\mu)=-1791+359\mu-27\mu^{2}+0.9\mu^{3}+0.011\mu^{4}$\\
%Hom, $\tau=10^5, V=200$\kms & $FWHM(\mu)=-51882.9+12932.5\mu-54.58\mu^{2}-187.22\mu^{3}+11.66\mu^{4}$ \\
%Hom, $\tau=10^6, V=200$\kms & $FWHM(\mu)=-7154649+2413478\mu-305268.4\mu^{2}+17158.92\mu^{3}-361.64\mu^{4}$ \\
%Hom, $\tau=10^7, V=200$\kms & $FWHM(\mu)=-51074+9912\mu-721\mu^{2}+23.29\mu^{3}-0.28\mu^{4}$ \\
%\hline
%\end{tabular}
%\caption{Fits for EW models} 
%\label{table:fits}
%\end{center}
%\end{table*}

\begin{table}
\begin{center}
\begin{tabular}{ccc}\hline          
Velocity (Km/s) & Maximum 1 position & Maximum 2 position \\ \hline
50 & -16.2695 & 16.23705 \\ 
100 & -15.66496 & 15.33504 \\ 
200 & -16.93149 & 14.56851 \\ 
300 & -13.40048 & 16.09952 \\ \hline 
\end{tabular} 
\caption{Optical Depth $\tau= 10^{7}$, Central Distribution}
\end{center}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}{ccc}\hline 
Velocity (km/s) & Maximum 1 position & Maximum 2 position\\ \hline
50 & -7.46286 & 6.53714 \\ 
100 & -7.53357 & 6.96643 \\ 
200 & -8.17453 & 7.32547 \\ 
300 & -6.81487 & 6.18513 \\ \hline 
\end{tabular} 
\end{center}
\caption{Optical Depth $\tau = 10^{6}$, Central Distribution}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}{ccc}\hline 
Velocity(Km/s) & Maximum 1 position & Maximum 2 position\\ \hline
50 & -4.33708 & 3.66292 \\ 
100 & -4.27326 & 3.72674 \\ 
200 & -3.7737 & 3.7263   \\ 
300 & -3.84903 & 4.15097 \\ \hline 
\end{tabular}
\caption{Optical Depth $\tau=10^{5}$, Central distribution} 
\end{center}
\end{table}

Line width\\

\begin{table}
\begin{center}
\begin{tabular}{ccc}\hline 
Velocity(Km/s) & FWHM & $\theta$\\ \hline 
50 & 12.62 & $0^{o}$ \\ 
50 & 12.72 & $45^{o}$ \\ 
50 & 12.93 & $90^{o}$ \\ 
100 & 14.07 & $0^{o}$ \\ 
100 & 14.48 & $45^{o}$ \\ 
100 & 15.00 & $90^{o}$ \\ 
200 & 16.90 & $0^{o}$  \\ 
200 & 18.51 & $45^{o}$ \\ 
200 & 21.24 & $90^{o}$ \\ 
300 & $24.69^{*}$ & $0^{o}$ \\ 
300 & $25.79^{*}$ & $45^{o}$  \\ 
300 & $27.73^{*}$ & $90^{o}$ \\ \hline
\end{tabular}
\caption{Lines Widhts for a Central Distribution and $\tau=10^{7}$} 
\end{center}
\end{table}

Escape fraction\\

\begin{table}[H]
\begin{center}
\begin{tabular}{ccccc}\hline 
Model & Velocity (km/s) & $\theta$ & Dust $\sum (s)$& $\sum (s)$\\ \hline 
Homogeneous & 50 & $0^{o}$& 13293.06 &49939.53\\ 
Homogeneous & 50 & $45^{o}$& 13291.04 &50001.59\\ 
Homogeneous & 50 & $90^{o}$ & 13348.76 &49922.73\\ 
Homogeneous & 100 & $0^{o}$ & 15527.69 &50114.11\\ 
Homogeneous & 100 & $45^{o}$ & 15511.56 &49967.17\\ 
Homogeneous & 100 & $90^{o}$ & 15401.71 & 49833.65\\ 
Homogeneous & 200 & $0^{o}$  & 17830.85 & 50078.69\\ 
Homogeneous & 200 & $45^{o}$ & 17932.87 & 50064.42\\ 
Homogeneous & 200 & $90^{o}$ & 17830.85  & 49931.748\\ 
Homogeneous & 300 & $ 0^{o}$ & 18687.33 & 50048.33 \\ 
Homogeneous & 300 & $45^{o}$ & 18572.12& 49922.67\\ 
Homogeneous & 300 & $90^{o}$ & 18421.79 & 49979.37\\ \hline
\end{tabular}
\caption{Escape fraction for a Homogeneous Distribution and optical depth $10^{5}$.} 
\end{center}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}{ccccc} \hline 
Model & Velocity (km/s) & $\theta$ & Dust $\sum (s)$& $\sum (s)$ \\  \hline 
Central & 50 & $0^{o}$ & 4809.881 & 49917.069 \\ 
Central & 50 & $45^{o}$ & 4829.21 & 49811.79 \\ 
Central & 50 & $90^{o}$ & 4845.108 & 49853.039\\ 
Central & 100 & $0^{o}$ & 4809.665 & 49921.30 \\  
Central & 100 & $45^{o}$ & 4828.65 & 49820.13 \\ 
Central & 100 & $90^{o}$ & 4846.45 & 49854.0 \\ 
Central & 200 & $0^{o}$  & 4809.63 & 49917.64 \\ 
Central & 200 & $45^{o}$ & 4829.25 & 49818.49 \\ 
Central & 200 & $90^{o}$ & 4844.89 & 49856.66 \\ 
Central & 300 & $ 0^{o}$ & 4810.56 & 49922.98\\ 
Central & 300 & $45^{o}$ & 4831.16 & 49823.33\\ 
Central & 300 & $90^{o}$ & 4845.33 & 49858.48\\ \hline
\end{tabular}
\caption{Escape fraction for the central Distribution and optical depth $10^{5}$.} 
\end{center}
\end{table}

%ideas:
Keep in mind making a follow up paper on the results of Yamada et atl http://adsabs.harvard.edu/abs/2012ApJ...751...29Y on the line morphology at z=3.1. How many of these laes can be considered to have rotation features? In principle, all of then should have them. How to make the case convincing in the statistical sense?

>>>>

\title[Rotation in the Lyman-$\alpha$ line]{Rotation effects on the
  Lyman-$\alpha$ line morphology in distant galaxies}
\author[N. Garavito and J.E. Forero-Romero]{
\parbox[t]{\textwidth}{\raggedright 
  Nicolas Garavito-Camargo$^{1}$,
  Jaime E. Forero-Romero$^{1}$ and 
  Mark Dijkstra$^2$
}
\vspace*{6pt}\\
$^{1}$Departamento de F\'{i}sica, Universidad de los Andes, Cra. 1
No. 18A-10, Edificio Ip, Bogot\'a, Colombia \\
$^2$
}
\maketitle

\begin{abstract}
Rotation is present in the gas kinematics of galaxies up to the
highest redshifts. In this paper we present for the first time
radiative transfer calculations that show the impact of rotation on
the morphology of the Lyman $\alpha$ line. To this end we construct
simplified models where a galaxy is modeled as an homogeneous sphere
composed as an homogeneous mixture of dust and hydrogen at a constant
temperature. These spheres have a solid-body rotation with linear
velocities at the surface in the range $0-300$ \kms. We consider
radiation sources both in the center of the rotating cloud and also
homogeneously distributed around the sphere. We find that higher
rotational velocities increase the width of each peak in the outgoing
line profile while it also increases the amount of Lyman alpha photons
escaping in the line center. This trends makes that for high
rotational velocities and large Hydrogen optical depths the double
peak of the line tends to be erased an be replaced by a single peak the
lines center. This is more pronounced for radiation sources
homogeneously distributed. Concerning the escape fraction we find that
rotation does not have any effect, provided that all the sources are
centrally emitted. However in the case of homogeneously emitted
sources we measure an increase of about a factor of $2$ in the escape
fraction for higher rotational velocity values. 
Our work shows clearly that gas rotation has a non
negligible impact on the shape of the Lyman $\alpha$ line.
\end{abstract}
\begin{keywords}
galaxies: high-redshift - galaxies: star formation - line: formation
\end{keywords}

@@

Due to the resonant nature of the Lyman alpha line, gas kinematics
play an important role shaping its morphology.

There is an extensive literature studying the influence of
outflow/inflow configurations in the shape of the outgoing Lyman-alpha
line.

In this paper we study for the first time the impact of rotation on
the morphology of the Lyman $\alpha$ line. To isolate the effects of
rotation we focus on a simple system: the gas distribution is
spherical, with homogeneous density and the gas rotates as a solid
body. We base our work on two independent Monte Carlo based radiative
transfer codes CLARA \citep{CLARA} and XX \citep{DijkstraKramer}.

@@

In this paper we experess a photon's frequency in terms of the
dimensionless variable $x\equiv (\nu -\nu_a)/\Delta\nu_\alpha$, where
$\nu_{\alpha}=2.46\times 10^{15}$ Hz is the Ly$\alpha$ resonance
frequency,  $\Delta_{\alpha} \equiv
\nu_{\alpha}\sqrt{2kT/m_pc^2}\equiv \nu_av_{\rm th} $ is the doppler
broadening of the line which depends on the neutral gas temperature
$T$ scattering the radiation or equivalently the thermal velocity
$v_{\rm th}$ of the atoms.

\section{Modeling Bulk Gas Rotation}
\label{sec:implementation}

Describing the kinematics of gas rotation in all generality is a
complex task. There is great variation in the shape of the rotation
curve as observed in HI emission as a function of the distance to the
galaxy center. However there are two features that are observed very
often. First, in the central region the velocity increases
proportional to the radius following the behaviour in a body with
solid rotation. Second, beyond a certain radius the rotation curve
tends to flatten. Furthermore, a thorough observational account of gas
rotation in the redshifts of interest for the study of LAEs ($z>1.0$)
is still missing.

An ab-initio description of realistic rotation curves in simulations
depends on having access to the dynamic evoluction the mass components
in the galaxy: stars, gas and dark matter. Such level of realism is
extremely complex to achive, specially of one wants to get a
systematic description based on statistics of simulated objects.

Following the tradition of studies of Lyman$\alpha$ emitting systems,
we implement a model with a simplified geoemtry and gas
distribution. We assume that the gas is homogeneously distributed in a
sphere that rotates as a solid body with constant angular
velocity. This simple model will contain only one parameter: the
linear velocity at the sphere's surface, $V_{\rm max}$.

\subsection{The implementation}
 We define a cartesian coordinate
system with its origin at the center of the sphere and the rotation
axis to be the $z$-axis, the components in the bulk velocity field, $\vec{v}
= v_{x}\hat{i} + v_{y}\hat{j} + v_{z}\hat{k}$. in the gas can be written as

@@

\subsection{Grid of Simulated Models}
\label{sec:models}

@@

@@

@@

@@

@@

@@

@@

\subsection{The Case of Asymmetric Emission}

@@

\begin{table*}
\begin{center}
\begin{tabular}{cccc}\hline
Model & SSA22a 4980/80   & HS1549 4667/80 & HS1700 4018/90\\
\hline
\end{tabular}
\caption{
Fluxes for tree different narrow band filters.
} 
\label{table:NBfilters}
\end{center}
\end{table*}

@@

Concerning the escape fraction under rotational effects on the
Ly$\alpha$ emission line, we found that the escape fraction increase
in about $20\%-30\%$for the homogeneous sphere model. While rotational
effects are negligible for the central models and the escape fraction
remains constant. Also the observer viewing angle have no effect in
the escape fraction neither for the homogeneous and central
models. Complementing this analysis we study the average number of
scatterings $<N_{scatt}>$ that photons perform before escaping of the
cloud taking into account rotational effects. The main result here is
for the homogeneous models for which as velocity increase photons
escape with about $\sim 39\%$  less scatterings than in the static
case.

@@

%ideas:
Keep in mind making a follow up paper on the results of Yamada et atl
http://adsabs.harvard.edu/abs/2012ApJ...751...29Y on the line
morphology at z=3.1. How many of t hese laes can be considered to have
rotation features? In principle, all of then should have them. How to
make the case convincing in the statistical sense?
